A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-10 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
Added citations
Added references for the definition of Wigner, Kirkwood-Dirac, and SIC-POVVM representations.

Adhikary 2020 defines that:
"The state is non-classical
with respect to these observables if, even one of the
pseudo-probabilities in the scheme is negative."

That should be sufficient to support your claim that there is a focus on negative probabilities, but better references might exist. I started with Asthana 2021 and found references therein.

I properly added the citation of Hall for the entropic volume.

\textbf{Quasi-probability representations.} Another way to represent quantum ensembles is through the use of quasiprobability distributions, of which the Wigner function is probably the most widely known. The Wigner transform $W : \Ens \to L^1(\mathbb{R}^2, \mathbb{R}) $, defined as [CITATION NEEDED]
\begin{equation}
	W(\rho) = W_{\hat{\rho}} (q, p) = 2 \int_{-\infty}^{\infty} \langle x + y | \rho | x - y \rangle e^{-2 i p y / \hbar} \; \mathrm dy,
\end{equation}
takes a mixed state $\rho$ and returns a function of phase space that integrates to one, like a probability density, but can have small negative regions. Recall that an affine combination is a linear combination where all coefficients sum to one but are not necessarily non-negative. Therefore, in general, quasi-probability distributions can be understood as affine combinations. Remarkably, expectation values of polynomial of position and momentum in symmetric ordering, coincide with expectations over the Wigner function. For example
\begin{equation}
	\tr \left( \frac{1}{2}[XP+PX] \rho \right) = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} xp \; W_{\rho} (q, p) dq dp \; .
\end{equation}
Moreover, the transform is an affine map: $W(p \rho_1 + \bar{p} \rho_2 ) = p W(\rho_1) + \bar{p} W(\rho_2)$.

In some problems, especially in quantum optics, normal ordering and anti-normal ordering are more useful. The Glauber P representation and the Husimi Q representation work like the Wigner function except that the expectations of polynomials recover the respective order. In other circumstances, quasi-probability distributions over observables different from position and momentum are more useful. For example, the Kirkwood-Dirac is used to represent two-time correlation functions over observables defined at different times. Quasi-probability representations are also used over discrete elements that may not even correspond to eigenstates of a particular operator. For example, the SIC-POVM represents quantum states as affine combinations of pure states $P_i$ that are ``equally distant'' from each other, which means that no two elements are orthogonal to each other. Each ensemble $\rho$ can then be represented as
\begin{equation}
	\rho = \sum_{n=1}^{d^2} \lambda_i P_i = \sum_{n=1}^{d^2} \left[ (d^2 + 1) p_i - \frac{1}{d} \right] P_i \; ,
\end{equation}
where $p_i = \tr \left(P_i \rho\right)$. [CITATION NEEDED]

@@

\textbf{Not about negative probability.} The inclusion of negative probability is often advertised as the main difference between classical probability and quasi-probability. [CITATION NEEDED] Sometimes these negative probabilities are justified as arising from constraints on probability distributions defined on different contexts. This is a mischaracterization. For example, the Husimi Q representation provides us with a non-negative quasi-probability function. That is, the Husimi Q is a non-negative density distribution that integrates to one, yet it is not a classical probability distribution.\footnote{Since the Husimi Q is defined as $Q_{\rho} ( \alpha ) = \frac{1}{\pi} \langle \alpha | \rho | \alpha \rangle$, it is proportional to the probability of measuring a coherent (i.e. Gaussian state) $| \alpha \rangle$, which is non-negative. } Similarly, a SIC-POVM characterizes the Bloch ball by four pure states that form a tetrahedron inscribed in the ball. Since we can also inscribe the Bloch ball inside a tetrahedron, we can represent all states of a two-state quantum system by a convex combination of the vertices, leading to a non-negative quasi-probability distribution over four elements.

@@

We now want to introduce a measure theoretic representation of the entropy that generalizes the notion of count of states. Since entropy characterizes the variability of the preparations of the ensemble, greater variability means the ensemble is spread over more cases. Therefore the count of distinguishable states must be a monotonic function of the entropy. It has been shown[cite Hall] under very general conditions that the exponential of the entropy is the only function to have the appropriate features. One required feature, that we prove in general, is that the exponential of the entropy is subadditive (i.e. the count of states associated to a mixture cannot exceed the combined count of states associated to both mixtures) and it is additive for orthogonal ensembles (i.e. no overlap in possible instances) in a specific linear combination.

Given a Borel set of ensembles $A \subseteq \Ens$, we define the \textbf{state capacity} of $A$ to be
\begin{equation}
\capacity(A) = \sup(2^{S(\hull(A))}\cup\{0\}).
\end{equation}
This returns the exponential of the biggest entropy achievable with a mixture of elements of $A$.

One can then show that the fraction capacity $\capacity_{\ens} : \Sigma_{\Ens} \to [0,\infty]$ is a set function that satisfies the following:
\begin{enumerate}
	\item non-negative - $\capacity(A)\geq 0$
	\item monotone - $A \subseteq B \implies \capacity(A) \leq \capacity(B)$
	\item sub-additive - $\capacity(A \cup B) \leq \capacity(A) + \capacity(B)$
	\item additive over orthogonal sets - $\capacity(A \cup B) = \capacity(A) + \capacity(B)$ if $A \ortho B$
	\item continuous from below and above - $\capacity(\lim\limits_{i \to \infty} A_i) = \lim\limits_{i \to \infty} \capacity(A_i)$ for any increasing or decreasing sequence $\{A_i\}$.
\end{enumerate}

@@

Again we note that the non-additive measure we recover is very well-behaved, and connects directly with important quantities. When Choquet theory applies, state capacity can be extended to the extreme points. In the classical case, disjoint sets of extreme points will be orthogonal, meaning that the state capacity there restricted is an additive measure. In the discrete case, every point is associated to zero entropy, and since $2^0=1$, the state capacity reduces to the counting measure. In the continuous case, gives a subset of phase space $A$, the maximum entropy is reached by the uniform distribution $\rho_A$, which means we recover $\log_2 \capacity(A) = S(\rho_A)$, the fundamental postulate of statistical mechanics. Similarly, in quantum mechanics, given an orthogonal subspace $A$, the maximum entropy is recovered by the maximally mixed state, which means $\capacity(A) = 2^{\dim(A)}$.

The state capacity, then, recovers the number of distinguishable states. In classical mechanics, all states are perfectly distinguishable. In quantum mechanics, only orthogonal state are distinguishable, and therefore the state capacity recovers the number of orthogonal states.

>>>>

\textbf{Quasi-probability representations.} Another way to represent quantum ensembles is through the use of quasiprobability distributions, of which the Wigner function is probably the most widely known. The Wigner transform $W : \Ens \to L^1(\mathbb{R}^2, \mathbb{R}) $, defined as \cite[eq.~11.8-3a]{mandel1995}
\begin{equation}
	W(\rho) = W_{\hat{\rho}} (q, p) = 2 \int_{-\infty}^{\infty} \langle x + y | \rho | x - y \rangle e^{-2 i p y / \hbar} \; \mathrm dy,
\end{equation}
takes a mixed state $\rho$ and returns a function of phase space that integrates to one, like a probability density, but can have small negative regions. Recall that an affine combination is a linear combination where all coefficients sum to one but are not necessarily non-negative. Therefore, in general, quasi-probability distributions can be understood as affine combinations. Remarkably, expectation values of polynomial of position and momentum in symmetric ordering, coincide with expectations over the Wigner function. For example
\begin{equation}
	\tr \left( \frac{1}{2}[XP+PX] \rho \right) = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} xp \; W_{\rho} (q, p) dq dp \; .
\end{equation}
Moreover, the transform is an affine map: $W(p \rho_1 + \bar{p} \rho_2 ) = p W(\rho_1) + \bar{p} W(\rho_2)$.

In some problems, especially in quantum optics, normal ordering and anti-normal ordering are more useful. The Glauber P representation and the Husimi Q representation work like the Wigner function except that the expectations of polynomials recover the respective order. In other circumstances, quasi-probability distributions over observables different from position and momentum are more useful. For example, the Kirkwood-Dirac is used to represent two-time correlation functions over observables defined at different times \cite{lostaglio2023}. Quasi-probability representations are also used over discrete elements that may not even correspond to eigenstates of a particular operator. For example, the SIC-POVM represents quantum states as affine combinations of pure states $P_i$ that are ``equally distant'' from each other, which means that no two elements are orthogonal to each other. Each ensemble $\rho$ can then be represented as \cite[eqs. 28-29]{fuchs2004}
\begin{equation}
	\rho = \sum_{n=1}^{d^2} \lambda_i P_i = \sum_{n=1}^{d^2} \left[ (d^2 + 1) p_i - \frac{1}{d} \right] P_i \; ,
\end{equation}
where $p_i = \tr \left(P_i \rho\right)$.

@@

\textbf{Not about negative probability.} The inclusion of negative probability is often advertised as the main difference between classical probability and quasi-probability \cite{adhikary2020}. Sometimes these negative probabilities are justified as arising from constraints on probability distributions defined on different contexts. This is a mischaracterization. For example, the Husimi Q representation provides us with a non-negative quasi-probability function. That is, the Husimi Q is a non-negative density distribution that integrates to one, yet it is not a classical probability distribution.\footnote{Since the Husimi Q is defined as $Q_{\rho} ( \alpha ) = \frac{1}{\pi} \langle \alpha | \rho | \alpha \rangle$, it is proportional to the probability of measuring a coherent (i.e. Gaussian state) $| \alpha \rangle$, which is non-negative. } Similarly, a SIC-POVM characterizes the Bloch ball by four pure states that form a tetrahedron inscribed in the ball. Since we can also inscribe the Bloch ball inside a tetrahedron, we can represent all states of a two-state quantum system by a convex combination of the vertices, leading to a non-negative quasi-probability distribution over four elements.

@@

\textbf{Counting states.} We now want to introduce a measure theoretic representation of the entropy that generalizes the notion of count of states. Since entropy characterizes the variability of the preparations of the ensemble, greater variability means the ensemble is spread over more cases. Therefore the count of distinguishable states must be a monotonic function of the entropy. Given that in statistical mechanics the entropy is the logarithm of the count of states, we expect the count of states to be the exponential of the entropy. Therefore one can argue that the proper way to measure the spread of an ensemble is through the exponential of the entropy, also known as entropic volume.\footnote{It has already been shown in \cite{hall1999} that the exponential of the entropy is the only function that can be used to define a volume.} However, the count of states, like the probability measure, is typically defined on the sample space $\Omega$, which can only work in the classical case. The entropy volume is a function of the ensemble space, so it is not a measure. Let's see everything fits together.

To build an intuition, consider the classical case. For every a region $U$ of phase space, we have a set of possible probability distributions that have support $U$. This is a convex set since every mixture of distributions with support $U$ has support $U$. Note that the uniform distribution $\rho_U$, if it exists, is the one with the greatest entropy, which will correspond to $S(\rho_U) = \log \mu(U)$. If $U$ is not compact, then there is no upper limit to the entropy and $\mu(U) = +\infty$. Now, let us transfer this intuition on the ensemble space $\ens$. If we take a set $A$, we can ask: what is the highest entropy reachable by a mixture of elements of $A$? If $A$ happens to be the set of ensembles with support $U$, then the supremum of the entropy is $\log \mu(U)$, which works also in the unbounded case. But the setup now is more general: $A$ can be any set of ensembles in any ensemble space.

We still need to make sure that this definition still makes sense physically. If the entropic volume measure the spread of an ensemble, during a mixture the spread can at most sum. Remarkably, the bounds on the entropy are such that the exponential of the entropy is subadditive. That is, $2^{S(p\ens[a]+\bar{p}\ens[b])} \leq 2^{S(\ens[a])} + 2^{S(\ens[b])}$ for any ensemble space.\footnote{It is an open question whether the exponential is the only monotonic function of the entropy that is subadditive.} Therefore this is indeed a general way to define the spread of ensembles and therefore the count of distinguishable states.

\textbf{State capacity.} Given a Borel set of ensembles $A \subseteq \Ens$, we define the \textbf{state capacity} of $A$ to be
\begin{equation}
\capacity(A) = \sup(2^{S(\hull(A))}\cup\{0\}).
\end{equation}
This returns the upper bound to the exponential of entropy achievable with a mixture of elements of $A$.

One can then show that the fraction capacity $\capacity : \Sigma_{\Ens} \to [0,\infty]$ is a set function that satisfies the following:
\begin{enumerate}
	\item non-negative - $\capacity(A)\geq 0$
	\item monotone - $A \subseteq B \implies \capacity(A) \leq \capacity(B)$
	\item sub-additive - $\capacity(A \cup B) \leq \capacity(A) + \capacity(B)$
	\item additive over orthogonal sets - $\capacity(A \cup B) = \capacity(A) + \capacity(B)$ if $A \ortho B$
	\item continuous from below and above - $\capacity(\lim\limits_{i \to \infty} A_i) = \lim\limits_{i \to \infty} \capacity(A_i)$ for any increasing or decreasing sequence $\{A_i\}$.
\end{enumerate}

@@

Again we note that the non-additive measure we recover is very well-behaved, and connects directly with important quantities. When Choquet theory applies, state capacity can be extended to the extreme points. In the classical case, disjoint sets of extreme points will be orthogonal, meaning that the state capacity there restricted is an additive measure. In the discrete case, every point is associated to zero entropy, and since $2^0=1$, the state capacity reduces to the counting measure. In the continuous case, gives a subset of phase space $U$ and $A_U$ the set of distributions with support within $U$, the maximum entropy is reached by the uniform distribution $\rho_U \in A_U$, which means we recover $\log_2 \capacity(A_U) = S(\rho_U)$, the fundamental postulate of statistical mechanics. Similarly, in quantum mechanics, given an orthogonal subspace $A$, the maximum entropy is recovered by the maximally mixed state, which means $\capacity(A) = 2^{\dim(A)}$.

The state capacity, then, recovers the number of distinguishable states. In classical mechanics, all states are perfectly distinguishable, the measure is additive. In quantum mechanics, only orthogonal state are distinguishable, therefore the measure is non-additive over pure states. We see again how the lack of single decomposition of ensembles over pure states is responsible for the failure of the additivity of the measure.
